An Integral Projection Modeling Approach to Understanding Demographic Effects of Multispecies Mutualisms

### 2024-02-22 Manuscript/ant_cactus_AmNat.tex
Manuscript Updates

\newpage{}
\section*{Introduction}
Mutualisms are species interactions where all participants receive net benefits, leading to higher individual fitness and increased population growth rates. 
They are among the most widespread species interactions \citep{Bronstein1994,Chamberlain2014,Frederickson2013,Axelrod1981,Leigh2010} but can deteriorate into commensalism or parasitism under conditions that elevate costs or dampen benefits \citep{Rodriguez-Rodriguez2017,Song2020,Mandyam2014,Thrall2007, Bahia2022}.
Mutualisms are considered more context dependent than other species interactions \citep{Chamberlain2014,Frederickson2013}, meaning the magnitude and sign of interaction strength are often determined by environmental conditions and species' identities \cite{Noe1994,Leigh2010}.

@@

There are multiple mechanisms by which partner diversity can influence the net benefits accrued by a focal mutualist, mirroring the mechanisms by which, at a larger scale of organization, biodiversity can influence ecosystem function \cite{Yeung2006,Barrett2015,Ushio2020}. 
First, when there is a hierarchy of fitness effects (a consistent ranking of best to worst mutualists), a more diverse sample of the partner community may be more likely to include the best partner \cite{Frederickson2013}.
This can lead to an apparent benefit of diversity driven by a sampling effect \cite{Batstone2018}. 
However, if partner associations are mutually exclusive (a focal mutualist may engage with only one partner at a time), then partner diversity may impose opportunity costs, leading to negative effects of a diverse mutualist assemblage relative to exclusive association with the single best partner \citep{Miller2007}. 
Second, even within a single mutualist guild, the benefits conferred by alternative partner species can vary in type and not just degree \cite{Stachowicz2005,Bronstein2006,Stanton2003}. 
This can lead to a positive effect of partner diversity through complementarity of alternative functions \cite{Batstone2018}. 
Interference or synergies between partners can make their combined effect different than the expected from the sum of complementary functions \cite{Afkhami2014}. 
Third, partner species and \tom{herbivores}{notice that ``herbiviores'' does not fit here, at least not yet. you have not said anything about ants and plants, you are talking very generally. put yourself in the mind of a reader as you write, this will help ensure that ideas appear in an order that can be digested.} can have species-specific responses to environmental variation, either spatially \citep{Ollerton2006} or temporally \citep{Alarcon2008}. 
Multiple partners can therefore act as a 'portfolio' that stabilizes fitness benefits and protection across spatial or temporal heterogeneity, leading to positive effects of partner diversity through the portfolio effect \cite{Batstone2018,Lazaro2022,Horvitz1990}.

@@

This study examined the consequences of partner diversity in a food-for-protection mutualism between the tree cholla cactus (\textit{Cylindriopuntia imbricata}), a long-lived EFN-bearing plant, and multiple species of ant partners.
Previous studies have shown that herbivory by specialized insect herbivores negatively affects plant fitness \cite{Miller2009}, and ant defense reduces herbivore damage \cite{Miller2007}. 
Tree cholla are tended by two common ant species (\textit{Liometopum apiculatum} and \textit{Crematogaster opuntiae}) and several infrequent species, all of which are ground-nesting. 
These ant species locally co-occur but individual plants are typically tended by only one species that patrols the plant around-the-clock and maintains control of the plant's nectar resources for an entire growing season \citep{Ohm2014, Donald2022}. 
Switches between partner species, or between vacancy and ant occupancy, commonly occur from one growing season to the next \citep{Miller2007}. 
Prior experiments suggested a hierarchy of mutualist quality, with \textit{Liometopum apiculatum} providing strong fitness benefits through anti-herbivore defense and \textit{Crematogaster opuntiae} having net negative fitness effects because herbivore deterrence is outweighed by deterrence of pollinators \citep{Miller2007,Ohm2014}. 
However, those studies did not integrate the demographic effects of ant defense across the plant life cycle, nor did they account for inter-annual fluctuations in the \tom{herbivore populations}{I would be careful here -- unless you are using the herbivore counts I don't think you have data on this.}.

@@

\subsection*{Data Collection}
This study is based a long-term demographic data set spanning 2004 to 2023 at SEV-LTER. 
From 2004 to 2008, we censused 134 plants distributed across three spatial blocks. 
This initial census group was discontinued in 2009, when we established six 30 $\times$ 30-meter plots and tagged all tree cholla within those plots. 
Two additional 30 $\times$ 30-meter plots were added in 2011, and this group of eight plots has since been censused annually through 2023 (with the exception of 2020 due to the covid shutdown). 
For all plants, in May or early June of each year we recorded plant survival since the last survey and, for survivors, we recorded the height (cm), maximum crown width (cm), and crown width perpendicular to the maximum (cm).
Size measurements were used to calculate plant volume ($cm^3$) based on the volume of an elliptical cone. 
We recorded reproductive effort as counts of viable and aborted flowerbuds. 
We recorded the ant species present (or vacancy if no ants present).
Occurrences of more than one ant species on one plant were rare (\tom{}{quantify}), and for the purpose of this analysis we classified the plant as being occupied by the more abundant species. 
Plots were searched for new recruits each year, and these were added to the census.
\tom{In total, the data set included \# unique individuals and \# plant-year observations. }{update numbers}
These data were used to fit vital rate models (survival, growth, reproduction) as functions of plant size and ant occupancy state.

@@

Given the low frequency of ant species other than \textit{L. apiculatum} and \textit{C. opuntiae} (\tom{}{Quantify this.}) we combined observations of all other ants into an ``other'' category, such that our models included four possible ant states: vacant, \textit{L. apiculatum}, \textit{C. opuntiae}, and ``other''. 
\tom{The ``other'' category was made of unidentified ant species and ant species which occurred at relatively low frequencies compared to \textit{C. opuntiae} and \textit{L. apiculatum} (such as unknown species belonging to the genus \textit{Aphenogaster} and unknown species belonging to the genus \textit{Camponotus}).}{I think you can do better than this because we actually have the designated species in the data. I would report here the actual breakdown (percentages) of species in the ``other'' group. I expect Forelious will be a big chunk of it.}
Ant state is included as a predictor variable in sub-models where there are biologically realistic pathways through which ants could impact the outcome of that process. 
For example, ant partners defend cacti from herbivores, and prior experimental work indicates that ant tending can reduce vegetative tissue loss and floral abortion.
Therefore, ant state was included in sub-models for survival, growth, and flowerbud viability. 
In contrast, we have no reason to expect that ant tending can directly influence the probability of flowering and flowerbud production, independently of its influence on plant size. 
Therefore, these sub-models included plant size but not ant state as predictor variables.

@@

\begin{linenomath*}
	$$
	B^1_{t+1} = \delta \sum_{a=1}^{A} \int_L^U  \kappa(a') P(x';\pmb{\tau^P}) F(x';\pmb{\tau^F}) V(a;\pmb{\tau^V_{a}}) n(x,a)_{t} dx \\
	$$
	$$
	B^2_{t+1} =  (1 - \gamma_1)B^1_{t}\\
	$$
\end{linenomath*}

\noindent Functions \tom{$P(x';\pmb{\tau^P})$ and $F(x';\pmb{\tau^F})$}{These should be $x$ not $x'$ because we modeled flowering and fertility in year t based on size in year t.} give the probability of flowering and the number of flowerbuds produced, respectively, by plants of size $x'$ in year $t$. 
The proportion of flowerbuds that remain viable through fruit set ($V(a;\pmb{\tau^V_{a}})$) and the number of seeds per fruit (\tom{$\kappa(a')$}{Why is this $a'$? I think it should be $a$. More generally, I think you need to explain your use of primes and how readers should interpret them at the start of model exposition.}) are dependent on ant state $a$ but not size. 
The vector $\pmb{\tau}$ gives year-specific deviates (with mean zero) and appears in functions for which we can estimate temporal stochasticity from the long-term data; superscripts indicate the corresponding vital rate and subscripts indicate that \tom{deviates are specific to plants in ant state $a$}{Why does the tau vector for viability have the $a$ subscript but the others do not?}. 
Seed production is integrated over the size distribution, from the lower $L$ to upper $U$ size limits, and summed over all possible ant states ($A$) giving total seed production. 
Seeds are multiplied by the probability of seed dispersal and survival ($\delta$) to give the number of seeds that enter the one-year old seed bank. 
Plants can recruit out of the year-one seed bank with probability $\gamma_1$ or transition to the two-year seed bank with a probability of $1 - \gamma_1$. 
Seeds in the two-year seed bank are assumed to either germinate with probability $\gamma_2$ or die.

@@

\subsection*{Statistical modeling and parameter estimation}
We parameterized the IPM using a series of generalized linear mixed models (GLMMs) in a hierarchical Bayesian framework to serve as vital rate sub-models. 
Vital rate models included spatial and temporal random effects associated with plot and year variation, respectively, and included plant size ($log(cm^3)$; $x,x'$), ant partner state ($a,a'$), or both as fixed-effect predictor variables. 
In addition to vital rate models describing plant demographic performance, we also fit a sub-model to predict transition between ant states conditional on plant size and previous ant state. 
All models were fit using Stan and Rstan (\tom{}{Cite Stan and RStan.}). 
Unless otherwise mentioned, all models use vague priors.

\paragraph{Growth}
The growth sub-model ($G(x',x,a;\pmb{\tau^G_{a}})$) gives the probability of future size given the fixed effects of previous size $x$ and previous ant partner $a$ and random effects of plot $w$ and year $u$. 
We fit this model to size data at the end of the transition year $y^G$ using the location-scale parameterization of the student $t$ distribution, because in preliminary analyses we found that size transition data were more fat-tailed than a Gaussian distribution could accommodate. 
Specifically, the model was: 
$$y^G \sim Student T(\hat{nu},\hat{G},\hat{\sigma}) $$
$$\hat{G} = \beta_{0}^{G} + \beta_{1}^{G} \times x  + \beta_{2}^{G} \times a + \beta_{3}^{G} \times x \times a  + \beta_{4}^{G} \times x^2 + \beta_{5}^{G} \times a \times x^2 + u + w $$
$$\hat{\sigma} = \beta_{0}^{\sigma} + \beta_{1}^{\sigma} \times x $$
$$\hat{\nu} = \beta_{0}^{\nu} + \beta_{1}^{\nu} \times x $$
where $u \sim N(0,\sigma_{yr \times a}^{2})$ is the year random effect with year specific variance of $\sigma_{yr}$ with ant $a$ in year $t$, and $w \sim N(0,\sigma_{plot}^{2})$ is the plot random effect with plot specific variance of $\sigma_{plot}$. 
Ants are included as predictors here because ant partners defend cholla from herbivores, preventing the loss of limbs and therefore shrinking.

\paragraph{Survival}
The survival model ($S(a,x;\pmb{\tau_{a}^{S}})$) estimates the probability of survival $y^S$ from year $t$ to year $t+1$, with fixed effects of the previous size of the cholla $x$ and ant partner $a$ in year $t$ and random effects of plot $w$ and year $u$, using a Bernoulli distribution:
$$y^S \sim Bern(\hat{S})$$
$$logit(\hat{S} = \beta_{0}^{S}) + \beta_{1}^{S} \times x + \beta_{2}^{S} \times a + \beta_{3}^{S} \times x \times a + u + w$$
where $u \sim N(0,\sigma_{yr \times a}^{2})$ is the year random effect with year specific variance of $\sigma_{yr}$ and ant state $a$ in year $t$, and $w \sim N(0,\sigma_{plot}^{2})$ is the plot random effect with plot specific variance of $\sigma_{plot}$.
A Bernoulli distribution was chosen here because the only two possible outcomes are survival or death. 
Ants are included as predictors here because ant partners defend cholla from herbivores and predators, decreasing the likelihood of mortality due to either of these.

\paragraph{Reproduction}
The reproduction model ($P(x';\pmb{\tau^{P}})$) estimates the probability of reproducing $y^P$ in year $t+1$, with fixed effects for the size $x'$ in year $t+1$ and random effects of plot $w$ and year $u$, using a Bernoulli distribution:
$$y^P \sim Bern(\hat{P})$$
$$logit(\hat{P}) = \beta_{0}^{P} + \beta_{1}^{P} \times x' + u + w$$
where $u \sim N(0,\sigma_{yr}^{2})$ is the year random effect with year specific variance of $\sigma_{yr}$, and $w \sim N(0,\sigma_{plot}^{2})$ is the plot random effect with plot specific variance of $\sigma_{plot}$.

\paragraph{Flowers}
The total flowers model ($F(x';\pmb{\tau^{F}})$) estimates the total flowers produced by a plant $y^F$ in year $t+1$, with fixed effects of size $x'$ in year $t+1$ and random effects of plot $w$ and year $u$, using a Negative Binomial distribution:
$$y^{F} \sim 0 Truncated Negative Binom(\hat{F},\hat{\phi})$$
$$log(\hat{F}) = \beta_{0}^{F} + \beta_{1}^{F} \times x' + u + w$$
$$log(\hat{\phi}) = \beta_{0}^{\phi}$$
where $u \sim N(0,\sigma_{yr}^{2})$ is the year random effect with year specific variance of $\sigma_{yr}$, and $w \sim N(0,\sigma_{plot}^{2})$ is the plot random effect with plot specific variance of $\sigma_{plot}$.

\paragraph{Viability}
The viability model ($V(a;\pmb{\tau^{V}_{a}})$) estimates the proportion of flowers produced by a plant which are viable (not aborted) $y^V$ in year $t+1$, with fixed effects of the ant partner of the cactus $a$ in year $t$ and random effects of plot $w$ and year $u$, using a Binomial distribution:
$$y^{V} \sim Binom(\hat{V})$$
$$logit(\hat{V}) = \beta_{0}^{V} \times a + u + w$$
where $u \sim N(0,\sigma_{yr \times a}^{2})$ is the year random effect with year specific variance of $\sigma_{yr}$ and ant state $a$ in year $t$, and $w \sim N(0,\sigma_{plot}^{2})$ is the plot random effect with plot specific variance of $\sigma_{plot}$.
Ants are included as predictors here because they defend the cacti from seed predation which can lead to floral abortion.

\paragraph{Ant Transitions}
The ant transition rates model ($\epsilon(x,a,a';\pmb{\tau^{\epsilon}})$) estimates the probability of a cactus being visited by an ant partner $a'$ $y^{\epsilon}$, with fixed effects of the previous size of the cholla $x$  and the previous ant partner $a$  in year $t$ and random effects of plot $w$ and year $u$, using a Multinomial distribution: 
$$y^{\epsilon} ~ Multinomial(\hat{\epsilon})$$
$$logit(\epsilon) = \beta_{0}^{\epsilon} + \beta_{1}^{\epsilon} \times x + \beta_{2}^{\epsilon} \times a + u$$
where $u \sim N(0,\sigma_{yr}^{2})$ is the year random effect with year specific variance of $\sigma_{yr}$.
Ant partners are included as predictors here because partners may choose to return to the same cholla repeatedly or choose new ones, therefore the previous partner may be a good indicator of the next partner.

\paragraph{Recruit Size Distribution}
The recruit size model ($n(x,a')$) estimates the size distribution of all recruits $y^{\eta}$ from a given year $t+1$, with no fixed or random effects, using a Normal distribution: 
$$y^{\eta} ~\sim N(\hat{\eta},\hat{\sigma})$$
$$\hat{\eta} = \beta_{0}^{\eta}$$
where $\hat{\sigma}$ is estimated with a non-informative prior.

\paragraph{Germination}
With germination data from \tom{Miller et al., 2007}{CITE}, we fit two Bayesian generalized linear models for the probability of germinating from a seed in the first year ($\gamma_1$) or the second year ($\gamma_2$) in year $t+1$, with no fixed or random effects, using a Binomial distribution:
$$y^{\gamma_1} \sim Binomial(\hat{\gamma_1})$$
$$y^{\gamma_2} \sim Binomial(\hat{\gamma_2})$$
$$logit(\hat{\gamma_1}) = \beta_{0}^{\gamma_1}$$
$$logit(\hat{\gamma_2}) = \beta_{0}^{\gamma_2}$$

\paragraph{Pre-Census Survival}
With data collected in a \tom{2005-2006}{FIX}********** recruit census, we fit a Bayesian generalized linear model for the probability of a seedling surviving to May ($\delta$) of year $t+1$ (accounting for missed mortality events), with fixed effects of the previous size $x$ and random effects of the transect $m$, using a Bernoulli distribution: 
$$y^{\delta} ~ Bern(\hat{\delta})$$
$$logit(\hat{\delta}) = \beta_{0}^{\delta} + m$$
 where $m \sim N(0, \sigma_{transect}^2)$ is the random effect of transect where the recruited individual was analyzed for survival.

\paragraph{Seeds Per Flower}
With data from \tom{Miller 2007}{CITE}********, we fit a Bayesian generalized linear model for the number of seeds produced by every flower on a cholla ($\kappa(a')$) in year $t+1$ based on the ant partner $a'$ in year $t+1$, using a Negative Binomial distribution:
$$y^{\kappa} \sim 0 Truncated Negative Binomial(\hat{\kappa},\hat{\phi})$$
$$ \hat{\kappa } = \beta_{0}^{\kappa} \times a'$$
$$\hat{\phi} = \beta_{0}^{\phi}$$
Ant partners are included as predictors here because they reduce floral abortion rates and therefore may lead to higher numbers of seeds. 
\tom{}{This paragraph needs to address how we assigned seed number to the ``other'' category.}

\paragraph{MCMC Simulations}
To obtain posterior estimates of the demographic parameters, we fit models using Markov chain Monte Carlo (MCMC) simulations via STAN run through version 4.0.2 of R \cite{Rcite,Rstancite} 
For each model, we obtained 3 chains of 10,000 iterations, each with randomly chosen initial conditions. 
The first 1,500 iterations were discarded as burn-in to eliminate transience associated with initial conditions. 
We did not thin the chains, thus all samples were retained. 
To assess the convergence of our models we assessed between and within chain convergence, the resulting figures are included in supplemental documents. 
To assess the overall model fit we carried out posterior predictive checks to examine how well the fitted model can generate simulated data similar to the real data.
Large differences in the two indicate a poor model fit and can be assessed visually (figures included in supplemental documents). 
All estimated parameters are available in the data. 
Data and code for all vital rate models is included in the supplemental information.

\subsection*{IPM Analysis}

>>>>

\newpage{}
\section*{Introduction}
Mutualisms are species interactions where all participants receive net benefits, leading to higher individual fitness and increased population growth rates. 
They are widespread species interactions \citep{Bronstein1994,Chamberlain2014,Frederickson2013,Axelrod1981,Leigh2010} but can deteriorate into commensalism or parasitism under conditions that elevate costs or dampen benefits \citep{Rodriguez-Rodriguez2017,Song2020,Mandyam2014,Thrall2007, Bahia2022}.
Mutualisms are considered more context dependent than other species interactions \citep{Chamberlain2014,Frederickson2013}, meaning the magnitude and sign of interaction strength are often determined by environmental conditions and species' identities \citep{Noe1994,Leigh2010}.

@@

There are multiple mechanisms by which partner diversity can influence the net benefits accrued by a focal mutualist, mirroring the mechanisms by which, at a larger scale of organization, biodiversity can influence ecosystem function \cite{Yeung2006,Barrett2015,Ushio2020}. 
First, when there is a hierarchy of fitness effects (a consistent ranking of best to worst mutualists), a more diverse sample of the partner community may be more likely to include the best partner \cite{Frederickson2013}.
This can lead to an apparent benefit of diversity driven by a sampling effect \cite{Batstone2018}. 
However, if partner associations are mutually exclusive (a focal mutualist may engage with only one partner at a time), then partner diversity may impose opportunity costs, leading to negative effects of a diverse mutualist assemblage relative to exclusive association with the single best partner \citep{Miller2007}. 
Second, even within a single mutualist guild, the benefits conferred by alternative partner species can vary in type and not just degree \cite{Stachowicz2005,Bronstein2006,Stanton2003}. 
This can lead to a positive effect of partner diversity through complementarity of alternative functions \cite{Batstone2018}. 
Interference or synergies between partners can make their combined effect different than the expected from the sum of complementary functions \cite{Afkhami2014}. 
Third, partner species can have species-specific responses to environmental variation, either spatially \citep{Ollerton2006} or temporally \citep{Alarcon2008}. 
Multiple partners can therefore act as a 'portfolio' that stabilizes fitness benefits and protection across spatial or temporal heterogeneity, leading to positive effects of partner diversity through the portfolio effect \cite{Batstone2018,Lazaro2022,Horvitz1990}.

@@

This study examined the consequences of partner diversity in a food-for-protection mutualism between the tree cholla cactus (\textit{Cylindriopuntia imbricata}), a long-lived EFN-bearing plant, and multiple species of ant partners.
Previous studies have shown that herbivory by specialized insect herbivores negatively affects plant fitness \cite{Miller2009}, and ant defense reduces herbivore damage \cite{Miller2007}. 
Tree cholla are tended by two common ant species (\textit{Liometopum apiculatum} and \textit{Crematogaster opuntiae}) and several infrequent species, all of which are ground-nesting. 
These ant species locally co-occur but individual plants are typically tended by only one species that patrols the plant around-the-clock and maintains control of the plant's nectar resources for an entire growing season \citep{Ohm2014, Donald2022}. 
Switches between partner species, or between vacancy and ant occupancy, commonly occur from one growing season to the next \citep{Miller2007}. 
Prior experiments suggested a hierarchy of mutualist quality, with \textit{Liometopum apiculatum} providing strong fitness benefits through anti-herbivore defense and \textit{Crematogaster opuntiae} having net negative fitness effects because herbivore deterrence is outweighed by deterrence of pollinators \citep{Miller2007,Ohm2014}. 
However, those studies did not integrate the demographic effects of ant defense across the plant life cycle, nor did they account for inter-annual fluctuations in the herbivore populations (data on herbivory is included in supplementary documents).

@@

\subsection*{Data Collection}
This study is based a long-term demographic data set spanning 2004 to 2023 at SEV-LTER. 
From 2004 to 2008, we censused 134 plants distributed across three spatial blocks. 
This initial census group was discontinued in 2009, when we established six 30 $\times$ 30-meter plots and tagged all tree cholla within those plots. 
Two additional 30 $\times$ 30-meter plots were added in 2011, and this group of eight plots has since been censused annually through 2023 (with the exception of 2020 due to the covid shutdown). 
For all plants, in May or early June of each year we recorded plant survival since the last survey and, for survivors, we recorded the height (cm), maximum crown width (cm), and crown width perpendicular to the maximum (cm).
Size measurements were used to calculate plant volume ($cm^3$) based on the volume of an elliptical cone. 
We recorded reproductive effort as counts of viable and aborted flowerbuds. 
We recorded the ant species present (or vacancy if no ants present).
Occurrences of more than one ant species on one plant were rare (less than 5\% of observed ant species), and for the purpose of this analysis we classified the plant as being occupied by the more abundant species. 
Plots were searched for new recruits each year, and these were added to the census.
In total, the data set included 1141 unique individuals and 19 plant-year observations. 
These data were used to fit vital rate models (survival, growth, reproduction) as functions of plant size and ant occupancy state.

@@

Given the low frequency of ant species other than \textit{L. apiculatum} and \textit{C. opuntiae} (7.78\% of observed ants) we combined observations of all other ants into an ``other'' category, such that our models included four possible ant states: vacant, \textit{L. apiculatum}, \textit{C. opuntiae}, and ``other''. 
The ``other'' category was made of unidentified ant species (2.8\% of observed ants), unknown species belonging to the geni \textit{Aphenogaster} (0.4\%),  \textit{Camponotus} (0.9\%),  \textit{Tetramorium} (0.02\%), \textit{Brachymyrmer} (0.02\%), a honeypot ant (0.08\%), and \textit{Forelius pruinosus} (3.5\%).
Ant state is included as a predictor variable in sub-models where there are biologically realistic pathways through which ants could impact the outcome of that process. 
For example, ant partners defend cacti from herbivores, and prior experimental work indicates that ant tending can reduce vegetative tissue loss and floral abortion.
Therefore, ant state was included in sub-models for survival, growth, and flowerbud viability. 
In contrast, we have no reason to expect that ant tending can directly influence the probability of flowering and flowerbud production, independently of its influence on plant size. 
Therefore, these sub-models included plant size but not ant state as predictor variables.

@@

\begin{linenomath*}
	$$
	B^1_{t+1} = \delta \sum_{a=1}^{A} \int_L^U  \kappa(a') P(x;\pmb{\tau^P}) F(x;\pmb{\tau^F}) V(a;\pmb{\tau^V_{a}}) n(x,a)_{t} dx \\
	$$
	$$
	B^2_{t+1} =  (1 - \gamma_1)B^1_{t}\\
	$$
\end{linenomath*}

\noindent In these equations $x$ and $x'$ indicate the size of a plant in years $t$ and $t+1$ respectively, $a$ and $a'$ indicate the ant partner of a plant in years $t$ and $t+1$.
Functions $P(x;\pmb{\tau^P})$ and $F(x;\pmb{\tau^F})$ give the probability of flowering in year $t$ and the number of flowerbuds produced in year $t$, respectively, by plants of size $x$. 
The proportion of flowerbuds that remain viable through fruit set ($V(a;\pmb{\tau^V_{a}})$) and the number of seeds per fruit ($\kappa(a)$) are dependent on ant state $a$ but not size. 
The vector $\pmb{\tau}$ gives year-specific deviates (with mean zero) and appears in functions for which we can estimate temporal stochasticity from the long-term data; superscripts indicate the corresponding vital rate and subscripts indicate that deviates are specific to plants in ant state $a$ and year.
Seed production is integrated over the size distribution, from the lower $L$ to upper $U$ size limits, and summed over all possible ant states ($A=4$) giving total seed production. 
Seeds are multiplied by the probability of seed dispersal and survival ($\delta$) to give the number of seeds that enter the one-year old seed bank. 
Plants can recruit out of the year-one seed bank with probability $\gamma_1$ or transition to the two-year seed bank with a probability of $1 - \gamma_1$. 
Seeds in the two-year seed bank are assumed to either germinate with probability $\gamma_2$ or die.

@@

\subsection*{Statistical modeling and parameter estimation}
We parameterized the IPM using a series of generalized linear mixed models (GLMMs) in a hierarchical Bayesian framework to serve as vital rate sub-models. 
Vital rate models included spatial and temporal random effects associated with plot and year variation, respectively, and included plant size ($log(cm^3)$; $x,x'$), ant partner state ($a,a'$), or both as fixed-effect predictor variables. 
In addition to vital rate models describing plant demographic performance, we also fit a sub-model to predict transition between ant states conditional on plant size and previous ant state. 
All models were fit using R version 4.2.2 and Rstan package (\cite{Rstancite, Rcite}).
Unless otherwise mentioned, all used vague priors.

The vital rate models take the form of the function:
$$ f(\mu) = \beta_0 + \beta_1 x + ... + u + w $$
\noindent where $u \sim N(0,\sigma_{yr }^{2})$ is the year random effect with year specific deviates of $\sigma_{yr}$ (which parameterizes the $\tau$ vector), and $w \sim N(0,\sigma_{plot}^{2})$ is the plot random effect with plot specific variance of $\sigma_{plot}$. 
Some models deviate from this general model, which will be specified below.

\paragraph{Growth}
The growth sub-model ($G(x',x,a;\pmb{\tau^G_{a}})$) gives the probability of future size given the fixed effects of previous size $x$ and previous ant partner $a$ and random effects of plot $w$ and year $u$ which interacts with ant state $a$. 
We fit this model to the size in year $t+1$ ($y^G$) using the location-scale parameterization of the student $t$ distribution and the identity link function, because in preliminary analyses we found that size transition data were more fat-tailed than a Gaussian distribution could accommodate. 
$$y^G \sim Student T(\hat{\nu},\hat{G},\hat{\sigma}) $$
Student $t$ distributions are dependent on three parameters, $\hat{\nu}$,  $\hat{G}$, and $\hat{\sigma}$.
$\hat{G}$ is the mean of the distribution, and in our case is expressed as a second-order polynomial with ant-size interactions because  preliminary analysis found this was a better fit to slightly concave data. 
$\hat{\nu}$ is the shape parameter which determines how thin the tails of the distribution are.
$\hat{\sigma}$ is the scale parameter which determines how wide the distribution is. 
Both $\hat{\nu}$ and $\hat{\nu}$ are size dependent because preliminary analyses revealed that there was size dependence in both variance and kurtosis.
%$$\hat{G} = \beta_{0}^{G} + \beta_{1}^{G}  x  + \beta_{2}^{G}  a + \beta_{3}^{G}  x  a  + \beta_{4}^{G}  x^2 + \beta_{5}^{G}  a  x^2 + u + w $$
$$\hat{\sigma} = \beta_{0}^{\sigma} + \beta_{1}^{\sigma}  x $$
$$\hat{\nu} = \beta_{0}^{\nu} + \beta_{1}^{\nu}  x $$

\paragraph{Survival}
The survival model ($S(a,x;\pmb{\tau_{a}^{S}})$) estimates the probability of survival from year $t$ to year $t+1$, with fixed effects of the previous size of the cholla $x$ and ant partner $a$ in year $t$ and random effects of plot $w$ and year $u$ which interacts with ant state $a$.
We fit this model to the survival data $y^S$  using a Bernoulli distribution and the logit link function
$y^S \sim Bern(\hat{S})$.
%$$logit(\hat{S}) = \beta_{0}^{S} + \beta_{1}^{S}  x + \beta_{2}^{S}  a + \beta_{3}^{S}  x  a + u + w$$

\paragraph{Reproduction}
The reproduction model ($P(x';\pmb{\tau^{P}})$) estimates the probability of reproducing in year $t+1$, with fixed effects for the size $x'$ in year $t+1$ and random effects of plot $w$ and year $u$.
We fit this model to the reproductive data $y^R$ using a Bernoulli distribution and a logit link function:
$$y^P \sim Bern(\hat{P})$$
%$$logit(\hat{P} = \beta_{0}^{P} + \beta_{1}^{P} \times x' + u + w)$$

\paragraph{No. Flowers}
The total flowers model ($F(x';\pmb{\tau^{F}})$) estimates the total flowers produced by a plant in year $t+1$, with fixed effects of size $x'$ in year $t+1$ and random effects of plot $w$ and year $u$. 
We fit this model to flowerbud count data $y^F$ using a zero-truncated negative binomial distribution with a log transformation:
$$y^{F} \sim 0 Truncated Negative Binom(\hat{F},\hat{\phi})$$
%$$log(\hat{F}) = \beta_{0}^{F} + \beta_{1}^{F} \times x' + u + w$$
$$log(\hat{\phi}) = \beta_{0}^{\phi}$$

\paragraph{Flowerbud viability}
The viability model ($V(a;\pmb{\tau^{V}_{a}})$) estimates theproportion of flowers produced by a plant which are viable (not aborted) in year $t+1$, with fixed effects of ant partner $a$ in year $t$ and random effects of plot $w$ and year $u$ which interacts with ant state $a$.
We fit this model to floral abortion data $y^V$ using a Binomial distribution and a logit link function:
$$y^{V} \sim Binom(\hat{V})$$
%$$logit(\hat{V}) = \beta_{0}^{V} \times a + u + w$$

\paragraph{Seeds Per Fruit}
With data\cite{Miller2006}, we fit a model for the number of seeds produced by every fruit on a cholla ($\kappa(a')$) in year $t+1$ based on the ant partner $a'$ in year $t+1$.
We fit this model to seed data $y^{\kappa}$ using a Negative Binomial distribution and the log link function: 
$$y^{\kappa} \sim  Negative Binomial(\hat{\kappa},\hat{\phi})$$
%$$ \hat{\kappa } = \beta_{0}^{\kappa} \times a'$$
$$\hat{\phi} = \beta_{0}^{\phi}$$
The data used for this model did not include data on ants in the ``other" category, so we used the data from vacant plants to parameterize seeds per flower for plants with ``other" ants in the IPM.

\paragraph{Ant Transitions}
The ant transition model ($\epsilon(x,a,a';\pmb{\tau^{\epsilon}})$) estimates the probability of a cactus being visited by an ant partner $a'$, with fixed effects of the previous size $x$  and the previous ant partner $a$  in year $t$ and random effects of plot $w$ and year $u$.
We fit this model to ant partner data from year $t+1$ $a'$ using a multinomial distribution with a logit link function: 
$$y^{\epsilon} ~ Multinomial(\hat{\epsilon})$$
%$$logit(\epsilon) = \beta_{0}^{\epsilon} + \beta_{1}^{\epsilon} \times x + \beta_{2}^{\epsilon} \times a + u$$

\paragraph{Recruit Size Distribution}
%The recruit size model ($n(x',a')$) estimates the size distribution of all recruits } from a given year $t+1$, with no fixed or random effects. 
We fit this model to recruit size data $y^{\eta}$ using a Normal distribution with the identity link function: 
$$y^{\eta} ~\sim N(\hat{\eta},\hat{\sigma})$$
%$$\hat{\eta} = \beta_{0}^{\eta}$$
where $\hat{\sigma}$ is estimated with a non-informative prior.

\paragraph{Germination}
With germination data \cite{Miller2007}, we fit two models for the probability of germinating from the first year seedbank ($\gamma_1$) or the second year seedbank ($\gamma_2$) in year $t+1$, with no fixed or random effects.
These models were fit to germination data $y^{\gamma_1}, y^{\gamma_2}$  using the binomial distribution with logit link functions:
$$y^{\gamma_1} \sim Binomial(\hat{\gamma_1})$$
$$y^{\gamma_2} \sim Binomial(\hat{\gamma_2})$$
%$$logit(\hat{\gamma_1}) = \beta_{0}^{\gamma_1}$$
%$$logit(\hat{\gamma_2}) = \beta_{0}^{\gamma_2}$$

\paragraph{Pre-Census Survival}
With recruit census data \cite{Miller2006}, we fit a model for the probability of a seedling (which germinates in early Fall) surviving to when we census in May ($\delta$) of year $t+1$ (accounting for missed mortality events), with fixed effects of the previous size $x$ and random effects of the transect $m$.
We fit this model to pre-census survival data $y^{\delta}$ using a Bernoulli distribution with a logit link function: 
$$y^{\delta} ~ Bern(\hat{\delta})$$
%$$logit(\hat{\delta}) = \beta_{0}^{\delta} + m$$
 where $m \sim N(0, \sigma_{transect}^2)$ is the random effect of transect where the recruited individual was analyzed for survival.

\paragraph{Parameter estimation}
To obtain posterior estimates of the demographic parameters, we fit models using Markov chain Monte Carlo (MCMC) simulations via STAN run through version 4.0.2 of R \cite{Rcite,Rstancite}. 
For each model, we obtained three chains of 10,000 iterations, each with randomly chosen initial conditions. 
The first 1,500 iterations were discarded. 
We did not thin the chains, thus all samples were retained following burn-in. 
We assessed parameter convergence between and within chains (Figures \ref{fig:Surv_post} -- \ref{fig:Pre_Surv_post} b). 
To assess the overall model fit we carried out posterior predictive checks to examine how well the fitted model can generate simulated data similar to the real data (Figures \ref{fig:Surv_post} -- \ref{fig:Pre_Surv_post} a).
All estimated parameters are available in Table \ref{tab:Params}.

\subsection*{IPM Analysis}
We constructed two stochastic IPMs in order to simulate ant combinations that don’t exist in real life with two different versions of environmental variation.
The first is referred to as the stochastic IPM, in which we include random annual variation in the model and allow each ant partner to respond distinctly to annual variation. 
The second is referred to as the stochastic null IPM, in which we include random annual variation in the model, but force each ant partner to respond in the same way to annual variation. 
By analyzing the stochastic model we can estimate the fitness of the cholla under these simulated combinations.
By comparing the stochastic model to the stochastic null model we can identify what mechanisms may explain differences in cholla fitness with different ant partners. 
In this section I first explain how a simple deterministic IPM is developed before diving into how we created our more complex stochastic and stochastic null models.
